---
title: "Assignment 4 Multivariate Models"
author: "James Strange"
date: "February 5, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(vegan)
library(dummies)
source('https://raw.githubusercontent.com/dmcglinn/quant_methods/gh-pages/scripts/utility_functions.R')
data(dune)
data(dune.env)
?dune
```

##########################################################

1.) Conduct an indirect ordination on the dune plant community. Specifically, visually examine a NMDS plot using the bray-curtis distance metric. Below is some code to help you develop a potential plot that emphasizes the role of the environmental variable "Moisture". Describe how you interpret the graphic. What is the goal of creating such a plot? Does this analysis suggest any interesting findings with respect to the dune vegetation?


```{r}
#Indirect ordination NMDS operation
dune_mds = metaMDS(dune)
#Plot results of NMDS ordination
plot(dune_mds, type='n', main = "Moisture NMDS")
color_vect = rev(terrain.colors(6))[-1]
text(dune_mds, 'sp', cex=.5)
points(dune_mds, 'sites', pch=19, 
       col=color_vect[dune.env$Moisture])
legend('topright', paste("Moisture =", 1:5, sep=''), 
        col=color_vect, pch=19)
```

The goal of creating the above NMDS graphic is to order the 20 sites based on similarities and differences in their species compositions and assign color code them according to the moisture index values with which they are associated. In other words, the graphic groups together sites with similar species compositions given the Moisture explanatory variable. Since the points that are color coded for Moisture index 1 are grouped relatively close together, these sites are more similar to each other in terms of species composition than they are to points representing plots of different mosture index values and therefore likely contain similar compositions of dune vegetation.   
Potential Hypotheses:
1) Sites with moisture index 1 have similar species compositions
2) The species Anthodor, Vicilath, Planlanc, Rumeacet, Trifprat, and Achimill are most closely associated with sites with Moisture index 1

###########################################################



2.) Carry out a direct ordination using CCA in order to test any potential hypotheses that you developed after examining the MDS plot. Specifically, carry out a test of the entire model (i.e., including all constrained axes) and also carry out tests at the scale of individual explanatory variables you included in your model if you included more than one variable. Plot your results. 


```{r}
#Dune CCA 
cca_dune = cca(dune ~ ., data=dune.env)
#Plot results
plot(cca_dune, type='n', scaling=3, main = "Dune Community Main CCA")
orditorp(cca_dune, display='sp', cex=0.5, scaling=3, col='blue')
text(cca_dune, display='bp', col='red')
plot(cca_dune, scaling=3, main = "Dune Community Main CCA")
```

Analysis of the above graphics suggests that the species Anthodor, Vicilath, Planlac, Rumeacet, Trifprat, and Achimill likely occurr in similar sites (i.e. sites with Moisture index 1 and the 'BF' Management treatment). 

```{r}
#Duna CCA stats
RsquareAdj(cca_dune, 100)
anova(cca_dune, permutations = 1000)
anova(cca_dune, by='margin', permutations = 1000)
##########################
#Tests for individual explanatory variables
##########################
#A1 explanatory variable
cca_dune_A1 = cca(dune ~ A1, data = dune.env)
RsquareAdj(cca_dune_A1, 100)
anova(cca_dune_A1, permutations = 1000)
anova(cca_dune_A1, by='margin', permutations = 1000)
#Moisture explanatory variable
cca_dune_Moisture = cca(dune ~ Moisture, data = dune.env)
RsquareAdj(cca_dune_Moisture, 100)
anova(cca_dune_Moisture, permutations = 1000)
anova(cca_dune_Moisture, by='margin', permutations = 1000)
#Management explanatory variable
cca_dune_Management = cca(dune ~ Management, data = dune.env)
RsquareAdj(cca_dune_Management, 100)
anova(cca_dune_Management, permutations = 1000)
anova(cca_dune_Management, by='margin', permutations = 1000)
#Use explanatory variable
cca_dune_Use = cca(dune ~ Use, data = dune.env)
RsquareAdj(cca_dune_Use, 100)
anova(cca_dune_Use, permutations = 1000)
anova(cca_dune_Use, by='margin', permutations = 1000)
#Manure explanatory variable
cca_dune_Manure = cca(dune ~ Manure, data = dune.env)
RsquareAdj(cca_dune_Manure, 100)
anova(cca_dune_Manure, permutations = 1000)
anova(cca_dune_Manure, by='margin', permutations = 1000)
############################
#Model Plots
plot(cca_dune_A1, scaling=3, main = "Dune Community A1 CCA")
plot(cca_dune_Moisture, scaling=3, main = "Dune Community Moisture CCA")
plot(cca_dune_Management, scaling=3, main = "Dune Community Management CCA")
plot(cca_dune_Use, scaling=3, main = "Dune Community Use CCA")
plot(cca_dune_Manure, scaling=3, main = "Dune Community Manure CCA")
```

Of the individual variable models, the Moisture model yields the highest R-squared value and the lowest p-value. The plots of the results indicate that Trifprat, Achimill, Vicilath, Rumeacet, Planlanc, and Anthodor are most losely associated with Moisture 1, as well as the 'BF' Management treatment. 

```{r}
#Model isolating Moisture and Management variables
cca_moisture.mgmt = cca(dune ~ Moisture + Management, data = dune.env)
RsquareAdj(cca_moisture.mgmt, 100)
anova(cca_moisture.mgmt, permutations = 1000)
anova(cca_moisture.mgmt, by='margin', permutations = 1000)
plot(cca_moisture.mgmt, scaling=3, main = "Dune Community Moisture/Management CCA")
```


###########################################################


3.) Do your two analyses agree with one another or complement one another or do these two analyses seem to be suggesting different take home messages? Which analysis do you find to be more useful?

The two analyses agree with one another because they both show the differences in environmental distribution of Anthodor, Vicilath, Planlac, Rumeacet, Trifprat, and Achimill to be closely tied to sites with Mosture 1 and Management BF. Specifically, the NMDS analysis shows that sites with Moisture 1 are relatively similar in terms of species composition and that Anthodor, Vicilath, Planlac, Rumeacet, Trifprat, and Achimill are closely associated with these sites. The CCA analyses provides further evidence that these species are associated with Moisture 1 sites, but also shows that a correlation to the Management variable is possible in that they are also tied to sites associated with the Biological Farming management treatment. 






