---
title: "Assignment 4 Multivariate Models"
author: "James Strange"
date: "February 5, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(vegan)
library(dummies)
source('https://raw.githubusercontent.com/dmcglinn/quant_methods/gh-pages/scripts/utility_functions.R')
data(dune)
data(dune.env)
?dune
```

##########################################################

1.) Conduct an indirect ordination on the dune plant community. Specifically, visually examine a NMDS plot using the bray-curtis distance metric. Below is some code to help you develop a potential plot that emphasizes the role of the environmental variable "Moisture". Describe how you interpret the graphic. What is the goal of creating such a plot? Does this analysis suggest any interesting findings with respect to the dune vegetation?


```{r}
#Indirect ordination NMDS operation
dune_mds = metaMDS(dune)
#Plot results of NMDS ordination
plot(dune_mds, type='n', main = "Moisture NMDS")
color_vect = rev(terrain.colors(6))[-1]
text(dune_mds, 'sp', cex=.5)
points(dune_mds, 'sites', pch=19, 
       col=color_vect[dune.env$Moisture])
legend('topright', paste("Moisture =", 1:5, sep=''), 
        col=color_vect, pch=19)
```

The goal of creating the above NMDS graphic is to order the sites based on species composition and color code them according to the moisture index levels with which they are associated. In other words, the graphic displays simmilarity/dissimilarity between sites based on species composition given the Moisture focus variable. Since the points that are color coded for Moisture index 1 are grouped relatively close together, these sites are more similar to each other than they are to points representing plots of different mosture index values and therefore likely contain similar compositions of dune vegetation species. In addition, points coding for sites associated with Moisture index 2 appear relatively close together and are therefore likely similar to each other in terms of species composition and are likely also similar to Moisture 1 sites in terms of species composition.  
Potential Hypotheses:
1) Sites with smaller Moisture index values (1-2) have similar species compositions
2) The species Trifprat, Achimill, Rumeacet, Vicilath, Lolipere, Anthodor, Planlanc, Poaprat, Bromhord, Poatriv, Elymrepe, and Cirsarve prefer habitats with smaller Moisture index values (1-2)

###########################################################



2.) Carry out a direct ordination using CCA in order to test any potential hypotheses that you developed after examining the MDS plot. Specifically, carry out a test of the entire model (i.e., including all constrained axes) and also carry out tests at the scale of individual explanatory variables you included in your model if you included more than one variable. Plot your results. 


```{r}
cca_dune = cca(dune ~ ., data=dune.env)
#Plot results
plot(cca_dune, type='n', scaling=3, main = "Dune Community CCA")
orditorp(cca_dune, display='sp', cex=0.5, scaling=3.5, col='blue')
text(cca_dune, display='bp', col='red')
plot(cca_dune, scaling=3)
```











3.) Do your two analyses agree with one another or complement one another or do these two analyses seem to be suggesting different take home messages? Which analysis do you find to be more useful?








