---
title: "Assignment 4 Multivariate Models"
author: "James Strange"
date: "February 5, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(vegan)
library(dummies)
source('https://raw.githubusercontent.com/dmcglinn/quant_methods/gh-pages/scripts/utility_functions.R')
data(dune)
data(dune.env)
?dune
```



1.) Conduct an indirect ordination on the dune plant community. Specifically, visually examine a NMDS plot using the bray-curtis distance metric. Below is some code to help you develop a potential plot that emphasizes the role of the environmental variable "Moisture". Describe how you interpret the graphic. What is the goal of creating such a plot? Does this analysis suggest any interesting findings with respect to the dune vegetation?


```{r}
#Inirect ordination NMDS
dune_mds = metaMDS(dune)
#Plot results of NMDS ordination
plot(dune_mds, type='n', main = "Moisture NMDS")
text(dune_mds, 'sp', cex=.5)
color_vect = rev(terrain.colors(6))[-1]
points(dune_mds, 'sites', pch=19, 
        col=color_vect[dune.env$Moisture])
legend('topright', paste("Moisture =", 1:5, sep=''), 
        col=color_vect, pch=19)
```

The goal of creating the above NMDS graphic is to order plant species distribution in the dune community based on site similarities in terms of the moisture level with which they are associated. It appears that sites exhibiting a Moisture index of 1 or 2 are roughly similar to each other in terms of their species distributions. 







2.) Carry out a direct ordination using CCA in order to test any potential hypotheses that you developed after examining the MDS plot. Specifically, carry out a test of the entire model (i.e., including all constrained axes) and also carry out tests at the scale of individual explanatory variables you included in your model if you included more than one variable. Plot your results. 


```{r}
cca_dune = cca(dune ~ ., data=dune.env)
anova(cca_dune, permutations = 999)
anova(cca_dune, by='margin', permutations = 999)
#Plot results
plot(cca_dune, type='n', scaling=1)
orditorp(cca_dune, display='sp', cex=0.5, scaling=1, col='blue')
text(cca_dune, display='bp', col='red')


```











3.) Do your two analyses agree with one another or complement one another or do these two analyses seem to be suggesting different take home messages? Which analysis do you find to be more useful?








